---
title: "Retrieving RWS pH data"
author: "Karline Soetaert, Willem Stolte, Sylvana Birchenough"
date: "16-6-2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# retrieving RWS pH data

Here we document the retrieval of the RWS (Rijkswaterstaat) data in the Dutch area of the North Sea. 

For this document, the data were retrieved at 17-6-2020.

A description (in dutch) of the RWS api is found here:
\url{https://rahuls01.github.io/?json#introduction}

The code to read the data was generated by Willem Stolte. It uses a package called rwsapi that can be installed from github:

```{r, eval = FALSE}
devtools::install_github("wstolte/rwsapi")
```

The scripts from this package ("rwsapi.r") can also be found in directory "../scripts".

First the metadata is read

```{r, eval = FALSE }
library(tidyverse)
library(lubridate)
library(rwsapi)
library(rlist)
library(httr)

# read the metadata set
  metadata <- rws_metadata()
  parmeta <- metadata$content$AquoMetadataLijst %>% 
    rlist::list.flatten()%>% as_tibble() %>% rename_all(tolower)
  locmeta <- metadata$content$LocatieLijst %>% rename_all(tolower)
  mapping <- metadata$content$AquoMetadataLocatieLijst %>% rename_all(tolower)
  ddlMetadata <- parmeta %>% full_join(mapping) %>% full_join(locmeta)
```

Find the name of parameter "pH" - this will be something as "zuurgraad" in Dutch.

This name is in the used AQUO standard in a field called *grootheid* (Dutch for quantity).

```{r, eval = FALSE }

grootheid = "zuurgraad"

# find exact parameter description
myPar <- ddlMetadata %>%
  filter(grepl(grootheid, tolower(ddlMetadata$parameter_wat_omschrijving))) %>%
  distinct(parameter_wat_omschrijving) %>% unlist() %>% unname()
```

A function to retrieve the data is created

```{r, eval = FALSE }

getDDLdata <- function(parameter, location, firstYear, lastYear, metadata){
  myMetadata <- metadata %>%
    filter(parameter_wat_omschrijving == parameter, code == location) %>%
    rename(locatie.code = code)
  
  beginDatumTijd <- paste0(firstYear, "-01-01T00:00:00.000+01:00")
  eindDatumTijd <- paste0(lastYear, "-12-31T23:59:59.000+01:00")
  
  getlist <- rws_makeDDLapiList(mijnCatalogus = myMetadata, beginDatumTijd, 
                                eindDatumTijd, "OW")
  
  df <- rws_observations2(getlist[[1]])$content
  return(df)
}
```

Find locations where pH is measured. The dataset also contains fresh water locations, so we select the marine locations from marineregions.org (VLIZ)

```{r, eval = FALSE }
locs <- ddlMetadata %>% 
  filter(parameter_wat_omschrijving == myPar) %>%
  distinct(code, x,y, coordinatenstelsel)

require(sf)
locations <- locs %>%
  st_as_sf(coords = c("x", "y"), crs = 25831) %>%
  st_transform(4326)   #  transform to wgs84

# Marine regions from marineregions.org (VLIZ)
layerurl <- "http://geo.vliz.be/geoserver/MarineRegions/ows?service=WFS&version=1.0.0&request=GetFeature&typeName=MarineRegions:eez_iho_union_v2&outputFormat=application/json"
marineRegions <- sf::st_read(layerurl)

# play a bit with the buffer (decimial degrees). 
# if zero, many freshwater stations are also selected.
marineLocs <- locations %>% sf::st_intersection(st_buffer(marineRegions, -0.05))

myLocs <- marineLocs %>% 
  st_drop_geometry() %>%
  distinct(code) %>%
  unlist() %>% unname()
```

Now loop over function getDDLdata for all locs and write each data set as a separate csv file

```{r, eval = FALSE }
Prefix <- paste("../data/raw_data/", myPar, del = "")
for(ii in 1:length(myLocs)){
  df <- getDDLdata(myPar, myLocs[ii], firstYear , lastYear, ddlMetadata)
  write_delim(df, paste(myPar,myLocs[ii], ".csv"))
} 
```

